---
title: Cell type references datasets
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
- name: Jared M. Andrews
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
- name: Friederike Dündar
  affiliation: Applied Bioinformatics Core, Weill Cornell Medicine
- name: Daniel Bunis
  affiliation: Bakar Computational Health Sciences Institute, University of California San Francisco, San Francisco, CA
date: "Revised: May 15th, 2020"
output:
  BiocStyle::html_document:
    toc_float: true
package: CellTypeReferences
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Available references}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

The `r Biocpkg("CellTypeReferences")` package provides convenient access to several cell type reference datasets. 
Most of these references are derived from bulk RNA-seq or microarray data of cell populations
that (hopefully) consist of a pure cell type after sorting and/or culturing.
Each dataset contains a log-normalized expression matrix that is intended to be comparable 
to log-UMI counts from common single-cell protocols,
typcially after adjusting bulk RNA-seq data for differences in gene length.
It also provides three levels of cell type annotation in its column metadata:

- `label.main`, broad annotation that defines the major cell types.
This has few unique levels that allows for fast annotation but at low resolution.
- `label.fine`, fine-grained annotation that defines subtypes or states.
This has more unique levels that results in slower annotation but at much higher resolution.
- `label.ont`, fine-grained annotation mapped to the standard vocabulary in the [Cell Ontology](https://www.ebi.ac.uk/ols/ontologies/cl).
This enables synchronization of labels across references as well as dynamic adjustment of the resolution.

By default, gene annotation is returned in terms of gene symbols,
but they can be coerced to Ensembl annotation with `ensembl=TRUE` for more robust cross-referencing across studies.
More details for each dataset can be viewed on the corresponding help page for its retrieval function (e.g., `?ImmGenData`).

# General-purpose references

## Human primary cell atlas (HPCA)

@hpcaRef
  
```{r}
library(CellTypeReferences)
ref <- HumanPrimaryCellAtlasData()
```

```{r tabulate, echo=FALSE}
samples <- colData(ref)[,c("label.main", "label.fine","label.ont")]
samples <- as.data.frame(samples)
DT::datatable(unique(samples))
```

## Blueprint/ENCODE

Blueprint [@blueprintRef] and Encode [@encodeRef],

```{r}
ref <- BlueprintEncodeData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

## Mouse RNA-seq

A collection of mouse data sets downloaded from GEO [@Benayoun2019].

```{r}
ref <- MouseRNAseqData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

# Immune references

## Database of Immune Cell Expression (DICE)

The [Database for Immune Cell Expression(/eQTLs/Epigenomics)](https://dice-database.org) [@diceRef].

```{r}
ref <- DatabaseImmuneCellExpressionData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

## Novershtern hematopoietic data

Novershtern Hematopoietic Cell Data - [GSE24759](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE24759) - formerly known as Differentiation Map [@dmapRef].

```{r}
ref <- NovershternHematopoieticData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

## Monaco immune data

Monaco Immune Cell Data - [GSE107011](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE107011)  [@monaco_immuneRef].

```{r}
ref <- MonacoImmuneData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

## Immunological Genome Project (ImmGen)

The murine [ImmGen](http://www.immgen.org/) [@ImmGenRef]

```{r}
ref <- ImmGenData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

# Session information {-}

```{r}
sessionInfo()
```

# References {-}
