---
title: Cell type references datasets
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
- name: Jared M. Andrews
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
- name: Friederike Dündar
  affiliation: Applied Bioinformatics Core, Weill Cornell Medicine
- name: Daniel Bunis
  affiliation: Bakar Computational Health Sciences Institute, University of California San Francisco, San Francisco, CA
date: "Revised: May 15th, 2020"
output:
  BiocStyle::html_document:
    toc_float: true
package: CellTypeReferences
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Available references}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

The `r Biocpkg("CellTypeReferences")` package provides convenient access to several cell type reference datasets. 
Most of these references are derived from bulk RNA-seq or microarray data of cell populations
that (hopefully) consist of a pure cell type after sorting and/or culturing.
Each dataset contains a log-normalized expression matrix that is intended to be comparable 
to log-UMI counts from common single-cell protocols [@aran2019reference],
typcially after adjusting bulk RNA-seq data for differences in gene length.
It also provides three levels of cell type annotation in its column metadata:

- `label.main`, broad annotation that defines the major cell types.
This has few unique levels that allows for fast annotation but at low resolution.
- `label.fine`, fine-grained annotation that defines subtypes or states.
This has more unique levels that results in slower annotation but at much higher resolution.
- `label.ont`, fine-grained annotation mapped to the standard vocabulary in the [Cell Ontology](https://www.ebi.ac.uk/ols/ontologies/cl).
This enables synchronization of labels across references as well as dynamic adjustment of the resolution.

By default, gene annotation is returned in terms of gene symbols,
but they can be coerced to Ensembl annotation with `ensembl=TRUE` for more robust cross-referencing across studies.
More details for each dataset can be viewed on the corresponding help page for its retrieval function (e.g., `?ImmGenData`).

# General-purpose references

## Human primary cell atlas (HPCA)

The HPCA reference consists of publicly available microarray datasets derived from human primary cells [@hpcaRef].
Most of the labels refer to blood subpopulations but cell types from other tissues are also available.

```{r}
library(CellTypeReferences)
ref <- HumanPrimaryCellAtlasData()
```

<hr>

```{r tabulate, echo=FALSE}
samples <- colData(ref)[,c("label.main", "label.fine","label.ont")]
samples <- as.data.frame(samples)
DT::datatable(unique(samples))
```

## Blueprint/ENCODE

The Blueprint/ENCODE reference consists of bulk RNA-seq data for pure stroma and immune cells 
generated by Blueprint [@blueprintRef] and ENCODE projects [@encodeRef].

```{r}
ref <- BlueprintEncodeData()
```

<hr>

```{r, echo=FALSE, ref.label="tabulate"}
```

## Mouse RNA-seq

This reference consists of a collection of mouse bulk RNA-seq data sets downloaded from the gene expression omnibus [@Benayoun2019].
A variety of cell types are available, again mostly from blood but also covering several other tissues.

```{r}
ref <- MouseRNAseqData()
```

<hr>

```{r, echo=FALSE, ref.label="tabulate"}
```

# Immune references

## Immunological Genome Project (ImmGen)

The ImmGen reference consists of microarray profiles of pure mouse immune cells from 
the [project of the same name](http://www.immgen.org/) [@ImmGenRef].
This is currently the most highly resolved immune reference -
possibly overwhelmingly so, given the fine granularity of the fine labels.

```{r}
ref <- ImmGenData()
```

<hr>

```{r, echo=FALSE, ref.label="tabulate"}
```

## Database of Immune Cell Expression/eQTLs/Epigenomics (DICE)

The DICE reference consists of bulk RNA-seq samples of sorted cell populations
from the [project of the same name](https://dice-database.org) [@diceRef].

```{r}
ref <- DatabaseImmuneCellExpressionData()
```

<hr>

```{r, echo=FALSE, ref.label="tabulate"}
```

## Novershtern hematopoietic data

The Novershtern reference (previously known as Differentiation Map)
consists of microarray datasets for sorted hematopoietic cell populations
from [GSE24759](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE24759) [@dmapRef].

```{r}
ref <- NovershternHematopoieticData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

## Monaco immune data

The Monaco reference consists of bulk RNA-seq samples of sorted immune cell populations 
from [GSE107011](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE107011) [@monaco_immuneRef].

```{r}
ref <- MonacoImmuneData()
```

```{r, echo=FALSE, ref.label="tabulate"}
```

# Session information {-}

```{r}
sessionInfo()
```

# References {-}
